Elaboration with First-Class Implicit Function Types

### 2020-02-27 paper.tex
stuff

There is significant literature on type inference in the presence of first-class
polymorphic types, mainly in relation to GHC and ML-like languages; see
e.g.\ \cite{leijen2008hmf,leijen2009flexible,serrano2018guarded,vytiniotis2006boxy}.
The above issue in Agda is a specific instance of the challenges described in
the mentioned works. So far, none of the above solutions have landed in
production compilers, for reasons of complexity, fragility and interaction with
other language features. A recent GHC development \cite{serrano2020a} offers a
solution which is relatively simple, and which is likely to land in an official
GHC release. However, none of these solutions support dependent types, and they
also have other limitations which we would like to address.

@@

\subsection{Contributions}
\begin{itemize}
  \item We propose an elaboration algorithm which translates from a small
    Agda-like surface language to a small Martin-L\"of type theory extended with
    implicit function types, telescopes and \emph{strictly curried function
      types} with telescope domain. We use these extensions to accummulate
    information about implicit insertions. Our algorithm is a conservative
    extension of Norell's bidirectional elaborator for Agda
    \cite[Chapter~3]{norell07thesis}.
  \item The presented elaborator is generally more powerful than previous
    solutions for impredicative inference. In particular, it goes beyond prior
    art by supporting full dependent type theory and by being global,
    i.e.\ inference may consider the whole program and not just particular n-ary
    applications.
  \item The target theory of elaboration serves as a general platform for
    elaborating implicit function types: our concrete elaborator is a relatively
    simple one, and there is room to further develop it.
  \item We provide an executable implementation of the elaborator described in
    this paper.
  \item Our solution is also \emph{simple}: we implemented elaboration,
    evaluation and unification in about 670 lines of Haskell, of which 210 lines
    implement the novel enhancements on the top of Norell's basic elaborator.
\end{itemize}

@@

We do not present any \emph{completeness} result for elaboration in this
paper. For an example of what this would entail, in \cite{dunfield2013complete},
completeness means that whenever there is a way to fill in missing details in
the surface syntax, algorithmic typechecking \emph{always} finds it. In
ibid.\ this means figuring out domain types for $\lambda$-s and inserting all
implicit applications. However, our elaborator targets a far stronger theory,
and it is beyond our reach to succinctly characterize which annotations are
inferable in the surface language, and we do not know of any prior work which
accomplishes this for a comparably strong elaborator. Our experience in Agda is
that it is not tractable in general to figure out which arguments are inferable,
by looking at function types; often we have to run elaboration to see what
works.

We can still say something about the behavior of our elaborator. For this, we
consider a translation from core terms to surface terms, the obvious forgetful
translation, which maps core terms to surface counterparts. Now, this is an
``evil'' construction on core terms, since it does not preserve definitional
equality, but we shall only use this evil notion in the following statement.

@@

We shall utilize the extended core theory to implement smarter
elaboration. Recall from Section \ref{sec:issues} that the old elaborator makes
two kinds of unforced insertion choices:
\begin{enumerate}
\item $\echeck{t}{\Theta}{\Gamma}{A}$ does not insert an implicit $\lambda$ when $A$ is meta-headed.
\item $\einsert{\true}{A}$ does not insert an implicit application if $A$ is meta-headed.
\end{enumerate}
In the following, we shall only enhance $\lambda$-insertions. This allows a
simple implementation which only requires minimal changes to unification, and
which is already remarkably powerful. It seems that handling implicit
application insertions requires extending unification; we discuss this in
Section \ref{TODO}. First, we modify closing types and contextualization to take
advantage of telescopes.

@@

We provide an implementation of the elaborator from Section
\ref{sec:extending_elaboration}. It is a standalone Haskell program which reads
a surface expression from standard input, and outputs the result of elaboration,
or optionally the type or the normal form of the result.

It is implemented in about 1200 lines of Haskell. Of this, 673 lines constitute
the core syntax, evaluation, unification and elaboration. Of this 673 lines, 458
lines implement the basic elaborator of Section
\ref{sec:bidirectional_elaboration} and 215 lines implement the extended
elaborator. We give here implementation-specific details and describe
differences to the presentation in this paper.

@@

Metacontexts are a key point in the implementation. We avoid the tedious (and
inefficient) threading of composed metasubstitutions, instead we have a mutable
reference which stores the current metasubstitution. We have a ``forcing''
operation which computes a semantic value to a head normal form with respect to
the current metasubstitution. Hence, instead of constantly updating every term
and type by performing metasubstitution, we only force them whenever we need to
match on the shape of types, for example when we are inserting implicit
arguments based on types. In this paper we still stick with the ``threaded
metasubstitution'' presentation because it is more conventional, and also more
straightforward to formalize.

We use normalization-by-evaluation in Abel's style (see
e.g.\ \cite[Chapter~3]{abel2013normalization}), where semantic values use de
Bruijn levels, and the core syntax uses de Bruijn indices. This is practically
very favorable, because the evaluator never has to perform weakening on values.
The implementation of curried functions presents a bit of a complication,
because the computation rules are type-directed, so we have to annotate
applications and abstractions with telescopes (just as in our notation for the
core syntax).

We provide a fairly optimized implementation for constraints: we keep track of
``blocking'' metas for each constraint, which are relevant to checking the
constancy condition, and upon solving a meta we only review constraints which
were blocked on the meta.

@@

\subsubsection{Related works}
MLF \cite{le2014mlf} is an extension of System F with polymorphic subtyping
bounds, which supports strong inference for first class polymorphism. HML
\cite{leijen2009flexible}) is a simplified variant of MLF. Since these systems
diverge markedly from (fragments of) Martin-Löf type theory and rely critically
on subtyping, it is unclear how they could be extended to dependent type
theories. HMF \cite{leijen2008hmf} is a relatively simple system with
first-class polymorphism; however, it is also weaker than others in terms of
inferable annotations, and its more powerful variant (extended with n-ary
application handling) jumps in complexity.

There have been several attempts in the context of GHC. Boxy types
\cite{vytiniotis2006boxy} and FPH \cite{vytiniotis2008fph} were two early
iterations which suffered from complex specification, complex implementation or
fragility. More recent works are guarded impredicativity (GI)
\cite{serrano2018guarded} and quick look impredicativity (QL)
\cite{serrano2020a}. They both try to examine n-ary applications to find
metavariable instantiations which arise from occurrences guarded by rigid type
constructors. GI's use of \emph{generalization constraints} anticipates our use
of telescopes, but overall GI is also burdened by inessential complexity. QL
streamlines GI by eschewing constraints in favor of an eager preprocessing pass
on neutral expressions which finds polymorphic instantiations. QL also takes
advantage of bidirectional type propagation.  Although QL seems to be
practically the most favorable so far, it is far from being elegant; it does
multiple checking passes, and the restriction of quick looking to neutral
applications is rather ad-hoc.

@@

\subsubsection{Evaluation}
We borrow a very useful collection of inference benchmarks from the GI
\cite{serrano2018guarded} and QL \cite{serrano2020a} papers: see Figure
\ref{fig:benchmark_types} and Figure \ref{fig:benchmark}. We also include a
source file in the implementation which reproduces these results.  In ibid.\ a
comparison is presented between multiple systems, but here we only include
FCIF. The relative performance of FCIC here is easy to remember: it handles
exactly the same cases as QL (which is slightly more than what GI covers). We
mark some cases as ``Yes*'', where FCIF successfully infers a type, but since
FCIC does no let-generalization, the inferred types are not fully constrained
without extra contextual information.

@@

While in this benchmark FCIF matches QL, in other ways it is significantly
stronger. Some general observations on the relative strength of FCIF:
\begin{itemize}
  \item
        Most obviously, FCIF supports dependent type theory and prior solutions
        do not. It is also doubtful whether prior solutions scale to dependent
        types. In particular, QL's preprocessing step is not applicable, since in
        a dependently typed spine we cannot do anything with later arguments until
        previous arguments are actually elaborated, or else it has to be restricted
        to non-dependent spines.
  \item
        FCIF is global and does not consider neutral spines specially in any
        way. For an example for global inference, $\slet\,x : \_ =
        single\,id\,\sin\,$ $cons\,\{IdTy\}\,x\,nil$ works in FCIF but does not
        work in GI, QL and HMF. It also works in MLF and HML, which are also
        global in this sense. These work by immediately giving a principal type
        to $single\,id$ which includes a subtyping bound. In contrast, FCIF
        makes no promises about principal typing --- which is not feasible with
        dependent types --- but it can still infer a type for $id$ which can be
        later constrained to $IdTy$.
\end{itemize}

\emph{Limitations.} What FCIF definitely cannot do, is inferring an implicit
function type for a function argument. MLF is the only system which can do this,
in cases where the polymorphic argument is used only once. Consider the
following:
\[
  \slet\,f : IdTy \to Bool = \lambda\,\_.\,true\,\sin\,\lambda\,x.\,f\,x
  \]
FCIF will attempt to elaborate $f\,x$ to $f\,(\lambda\{A\}.\,x)$, but fails to
infer a type for $x$, because $A$ is not in the scope of $x$'s type. In the next
section we sketch an extension of FCIF which can handle this, and can also
handle $B2$ from Figure \ref{fig:benchmark}.

\section{Future Work}

\section{Conclusion}

%% -- let f : IdTy → Bool = λ x. true in
%% -- λ x. f x

>>>>

There is significant literature on type inference in the presence of first-class
polymorphic types, mainly in relation to GHC and ML-like languages; see
e.g.\ \cite{leijen2008hmf,leijen2009flexible,serrano2018guarded,vytiniotis2006boxy}.
The above issue in Agda is a specific instance of the challenges described in
the mentioned works. Currently none of the above solutions are supported in
production compilers, for reasons of complexity, fragility and interaction with
other language features. A recent GHC development \cite{serrano2020a} offers a
solution which is relatively simple, and which is likely to land in an official
GHC release. However, none of these works support dependent types, which is a
key point in our work.

@@

\subsection{Contributions}
\begin{itemize}
  \item We propose an elaboration algorithm which translates from a small
    Agda-like surface language to a small Martin-L\"of type theory extended with
    implicit function types, telescopes and \emph{strictly curried function
      types} with telescope domain. We use these extensions to accumulate
    information about implicit insertions. Our algorithm is based on Norell's
    bidirectional elaborator for Agda \cite[Chapter~3]{norell07thesis}.
  \item
    In the System F fragment, the presented elaborator is comparable or superior
    to previous solutions for impredicative inference. However, it also supports
    full dependent type theory. Our inference is also global, i.e.\ it can
    consider the whole program and not just particular n-ary applications.
  \item We provide an executable implementation of the elaborator described in
    this paper.
  \item Our solution is simple: we implemented elaboration,
    evaluation and unification in about 670 lines of Haskell, of which 210 lines
    implement the novel enhancements on the top of Norell's basic elaborator.
  \item The target theory of elaboration serves as a general platform for
    elaborating implicit function types: our concrete elaborator is a relatively
    simple one, and there is room to further develop it.
\end{itemize}

@@

We do not present any \emph{completeness} result for elaboration in this
paper. For an example of what this would entail, in \cite{dunfield2013complete}
completeness means that whenever there is a way to fill in missing details in
the surface syntax, algorithmic typechecking \emph{always} finds it. In
ibid.\ this means figuring out domain types for $\lambda$-s and inserting all
implicit applications. However, our elaborator targets a far stronger theory,
and it is beyond our reach to succinctly characterize which annotations are
inferable in the surface language, and we do not know of any prior work which
accomplishes this for a comparably strong elaborator. Our experience in Agda is
that it is not tractable in general to figure out which arguments are inferable,
by looking at function types, and often we have to run elaboration to see what
works.

We can still say something about the behavior of our elaborator. For this, we
consider a translation from core terms to surface terms, the evident forgetful
translation, which maps core terms to surface counterparts. Now, this is an
``evil'' construction on core terms, since it does not preserve definitional
equality, but we shall only use this evil notion in the following statement.

@@

We shall utilize the extended core theory to implement smarter
elaboration. Recall from Section \ref{sec:issues} that the old elaborator makes
two kinds of unforced insertion choices:
\begin{enumerate}
\item $\echeck{t}{\Theta}{\Gamma}{A}$ does not insert an implicit $\lambda$ when $A$ is meta-headed.
\item $\einsert{\true}{A}$ does not insert an implicit application if $A$ is meta-headed.
\end{enumerate}
In the following, we shall only enhance $\lambda$-insertions. This allows a
simple implementation which only requires minimal changes to unification, and
which is already remarkably powerful. It seems that handling implicit
application insertions requires extending unification; we discuss this in
Section \ref{sec:appinsert}. First, we modify closing types and
contextualization to take advantage of telescopes.

@@

We provide an implementation of the elaborator from Section
\ref{sec:extending_elaboration}. It is a standalone Haskell program which reads
a surface expression from standard input, and outputs the result of elaboration,
or optionally the type or the normal form of the result. It is implemented in
about 1200 lines of Haskell. Of this, 673 lines constitute the core syntax,
evaluation, unification and elaboration. Of this 673 lines, 458 lines implement
the basic elaborator of Section \ref{sec:bidirectional_elaboration} and 215
lines implement the extended elaborator.

@@

Metacontexts are a key point in the implementation. We avoid the tedious (and
inefficient) threading of composed metasubstitutions, instead we have a mutable
reference which stores the current metasubstitution. We have a ``forcing''
operation which computes a semantic value to a head normal form with respect to
the current metasubstitution. Hence, instead of constantly updating every term
and type by performing metasubstitution, we only force them whenever we need to
pattern match on the shape of types, for example when we are inserting implicit
arguments based on types. In this paper we still stick with the ``threaded
metasubstitution'' presentation because it is more conventional, and also more
straightforward to formalize.

We use normalization-by-evaluation in Abel's style (see
e.g.\ \cite[Chapter~3]{abel2013normalization}), where semantic values use de
Bruijn levels, and the core syntax uses de Bruijn indices. This is practically
very favorable, because the evaluator never has to perform weakening on values.
The implementation of curried functions presents a bit of a complication,
because the computation rules are type-directed, so we have to annotate
applications and abstractions with telescopes (just as in our notation for the
core syntax). We implement constraints in a fairly optimized way: we keep track
of relevant ``blocking'' metas for each constraint and upon solving a meta we
only review constraints which were blocked on the meta.

@@

\subsection{Related Works}
MLF \cite{le2014mlf} is an extension of System F with polymorphic subtyping
bounds, which supports strong inference for first class polymorphism. HML
\cite{leijen2009flexible}) is a simplified variant of MLF. Since these systems
diverge markedly from (fragments of) Martin-Löf type theory and rely critically
on subtyping, it is unclear how they could be extended to dependent type
theories. HMF \cite{leijen2008hmf} is a relatively simple system with
first-class polymorphism; however, it is also weaker than others in terms of
inferable annotations, and its more powerful variant (extended with n-ary
application handling) jumps in complexity.

There have been several attempts in the context of GHC. Boxy types
\cite{vytiniotis2006boxy} and FPH \cite{vytiniotis2008fph} were two early
iterations which suffered from complex specification, complex implementation or
fragility. More recent works are guarded impredicativity (GI)
\cite{serrano2018guarded} and quick look impredicativity (QL)
\cite{serrano2020a}. They both try to examine n-ary applications to find
metavariable instantiations which arise from occurrences guarded by rigid type
constructors. GI's use of \emph{generalization constraints} anticipates our use
of telescopes, but overall GI is also burdened by inessential complexity. QL
streamlines GI by eschewing constraints in favor of an eager preprocessing pass
on neutral expressions which finds polymorphic instantiations. QL also takes
advantage of bidirectional type propagation.  Although QL seems to be
practically the most favorable so far, it is far from being elegant: it rechecks
expressions multiple times, and the restriction of preprocessing to (nested)
neutral applications is rather ad-hoc.

@@

\subsection{Evaluation}
We borrow a very useful collection of inference benchmarks from the GI
\cite{serrano2018guarded} and QL \cite{serrano2020a} papers: see Figure
\ref{fig:benchmark_types} and Figure \ref{fig:benchmark}. We also include a
source file in the implementation which reproduces these results.  In ibid.\ a
comparison is presented between multiple systems, but here we only include
FCIF. The relative performance of FCIF here is easy to remember: it handles
exactly the same cases as QL (which is slightly more than what GI covers). We
mark some cases as ``Yes*'', where FCIF successfully infers a type, but since
FCIF does no let-generalization, the inferred types are not fully constrained
without extra contextual information.

@@

Some general comments on the comparison of FCIF to prior works. First, FCIF
supports dependent types and prior solutions do not. It is also unclear whether
prior solutions can scale to dependent types. In particular, QL's preprocessing
step is not applicable, since in a dependently typed spine we cannot do anything
with later arguments until previous arguments are actually elaborated, or else
it has to be restricted to non-dependent spines.

Secondly, FCIF supports global inference and does not handle neutral spines
specially. For an example for global inference, $\slet\,x : \_ =
single\,id\,\sin\,$ $cons\,\{IdTy\}\,x\,nil$ works in FCIF, MLF and HMF but does
not work in GI, QL and HMF. MLF and HML work here by immediately giving a
principal type to $single\,id$ which involves generalization with a subtyping
bound. In contrast, FCIF makes no promises about principal typing --- which is
not feasible with dependent types --- and does no generalization, but it can
still infer a type for $id$ which can be later constrained to $IdTy$.

\subsection{Inferring Polymorphic Types for Function Arguments}
\label{sec:appinsert}

It is apparent from the $B1$ and $B2$ cases on Figure \ref{fig:benchmark} that FCIF
cannot infer implicit function types for function arguments.  MLF is the only
system which can do this, in cases where the polymorphic argument is used only
once. Consider the following:
\[
  \slet\,f : IdTy \to Bool = \lambda\,\_.\,true\,\sin\,\lambda\,x.\,f\,x
  \]
FCIF will attempt to elaborate $f\,x$ to $f\,(\lambda\{A\}.\,x)$, but fails to
infer a type for $x$, because $A$ is not in the scope of $x$'s type. We sketch
an extension of FCIF with \emph{implicit application insertion}, which could
possibly handle this example. We note though that this is not yet fleshed out
and is a subject of future work.

\section{Future Work and Conclusion}
